---
title: "A Predicted Increase in the Intensity of Strong Atlantic Hurricanes: Validating Elsner et al. (2008)"
author: "James B. Elsner"
date: September 19, 2019
output: html_document
---

Science is validate through predictions. Using satellite derived wind speed estimates from tropical cyclones [Elsner et al. (2008)](http://myweb.fsu.edu/jelsner/PDF/Research/ElsnerKossinJagger2008.pdf) showed that the strongest hurricanes got stronger in the period 1981-2006. In particular, they found that 15% (85th quantile) of all named storms across the North Atlantic had wind speeds exceeding 48.9 m/s (Table 1) with an upward trend of .63 m/s/yr in this quantile wind speed. They related this increase to a corresponding rise in ocean temperatures consistent with [theory](https://www.nature.com/articles/nature03906). Since the oceans have continued to warm since that paper was published we would predict that the upward trend in the intensity of the strongest hurricanes to continue. 

Now considering all North Atlantic named storms (tropical storms and hurricanes) over the 13-year period 2007-2019 [data taken from Wikipedia] we note that 20% of them had wind speeds exceeding 48.9 m/s. An increase of 5 percentage points.
```{r}
Year <- c(2007, 2007, 2008, 2008, 2008, 2008, 2008, 2009, 2009, 2010, 2010, 2010, 2010, 2010,
          2011, 2011, 2011, 2011, 2012, 2012, 2014, 2014, 2015, 2015, 2016, 2016, 2016, 2016,
          2017, 2017, 2017, 2017, 2017, 2017, 2018, 2018, 2019, 2019)
Name <- c("Dean", "Felix", "Bertha", "Gustav", "Ike", "Omar", "Paloma", "Bill", "Fred",
          "Danielle", "Earl", "Igor", "Julia", "Karl", "Irene", "Katia", "Ophelia", "Rina",
          "Michael", "Sandy", "Edouard", "Gonzalo", "Danny", "Joaquin", "Gaston", "Matthew",
          "Nicole", "Otto", "Harvey", "Irma", "Jose", "Lee", "Maria", "Ophelia", "Florence",
          "Michael", "Dorian", "Humburto")
MPH <- c(175, 175, 125, 150, 145, 130, 140, 130, 120, 130, 145, 155, 140, 125, 120, 140, 140,
         115, 115, 115, 120, 145, 125, 155, 120, 165, 140, 115, 130, 180, 155, 115, 175, 115, 
         150, 160, 185, 115)
LMI.df <- data.frame(Year, Name, MPH, MPS = MPH * .44704)

NS <- c(15, 16, 9, 19, 19, 19, 14, 8, 11, 15, 17, 15, 10)
MH <- c(2, 5, 2, 5, 4, 2, 0, 2, 2, 4, 6, 2, 2)
Year <- 2007:2019

Counts.df <- data.frame(Year, NS, MH)
sum(MH)/sum(NS) * 100
```

```{r}
library(dplyr)
LMI.df %>%
  arrange(desc(MPS))
```

Create an ordered bar plot.
```{r}
LMI.df <- LMI.df %>%
  mutate(NAME = paste(Year, Name))

library(ggplot2)
library(forcats)

ggplot(data = LMI.df) +
  geom_segment(aes(x = reorder(NAME, MPS), y = 50,
                   xend = reorder(NAME, MPS), yend = MPS, 
                   color = MPS), size = 2.5) +
  coord_flip() + 
  scale_y_continuous(position = 'right') +
  xlab("") + ylab("Lifetime Highest Intensity (m/s)") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none")
```

Further we note that the 85th quantile wind speed has increased from 48.9 m/s to 53.1 m/s which is an increase of 4.2 m/s and which matches an extrapolation by taking the trend of .63 m/s/yr from Table 1 of Elsner et al. (2008) and multipling it by 7 (mid point of the 13-year period). [48.9 + .63 * 7 = 53.3 m/s].
```{r}
percentile <- (cumsum(table(LMI.df$MPS)) + (sum(NS) - sum(MH)))/sum(NS)
percentile.df <- as.data.frame(percentile)
percentile.df$MPS <- as.numeric(rownames(percentile.df))

approx(x = percentile.df$percentile, y = percentile.df$MPS, xout = .85)
```

Furthermore we note that only 5% of the storms exceeded 60.3 m/s over the earlier period (Table 1, 95th quantile) but that has increased to 10% (1 - .90) and the 95th quantile wind speed has increased from 60.3 m/s to 67.5 m/s consistent with extrapolating the trend [60.3 + .81 * 7 = 66 m/s].
```{r}
approx(x = percentile.df$MPS, y = percentile.df$percentile, xout = 60.3)
approx(x = percentile.df$percentile, y = percentile.df$MPS, xout = .95)
```

It is very hard to argue against this simple post publication analysis and it raises the question whether folks were negligent in ignoring this work when writing "state-of-the-knowledge" reports on hurricanes and climate change.