---
title: The Increasing Intensity of Strong Atlantic Hurricanes
subtitle: Validating Elsner et al. (2008)
author: James B. Elsner
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
bibliography: References.bib
editor_options: 
  chunk_output_type: console
---

Science gets validated through predictions. Using satellite derived wind speed estimates from tropical cyclones @ElsnerKossinJagger2008 showed that the strongest hurricanes were getting stronger through the period 1981-2006. In particular, they found that 15% (85th quantile) of all named storms across the North Atlantic had wind speeds exceeding 48.9 m/s (Table 1 in @ElsnerKossinJagger2008 ) with an upward trend of .63 m/s/yr in this quantile wind speed. They related this increase to a corresponding rise in ocean temperatures consistent with theory [@Emanuel1988]. The oceans have continued to warm since that paper was published so we would predict that the upward trend in the intensity of the strongest hurricanes has continued. 

To check this, consider all North Atlantic named storms (tropical storms and hurricanes) over the 13-year period 2007--2019 [data taken from Wikipedia]. `NS` and `MH` are the annual number of named storms and major hurricanes respectively.
```{r}
Year <- c(2007, 2007, 2008, 2008, 2008, 2008, 2008, 2009, 2009, 2010, 2010, 2010, 2010, 2010,
          2011, 2011, 2011, 2011, 2012, 2012, 2014, 2014, 2015, 2015, 2016, 2016, 2016, 2016,
          2017, 2017, 2017, 2017, 2017, 2017, 2018, 2018, 2019, 2019, 2019)
Name <- c("Dean", "Felix", "Bertha", "Gustav", "Ike", "Omar", "Paloma", "Bill", "Fred",
          "Danielle", "Earl", "Igor", "Julia", "Karl", "Irene", "Katia", "Ophelia", "Rina",
          "Michael", "Sandy", "Edouard", "Gonzalo", "Danny", "Joaquin", "Gaston", "Matthew",
          "Nicole", "Otto", "Harvey", "Irma", "Jose", "Lee", "Maria", "Ophelia", "Florence",
          "Michael", "Dorian", "Humburto", "Lorenzo")
MPH <- c(175, 175, 125, 150, 145, 130, 140, 130, 120, 130, 145, 155, 140, 125, 120, 140, 140,
         115, 115, 115, 120, 145, 125, 155, 120, 165, 140, 115, 130, 180, 155, 115, 175, 115, 
         150, 160, 185, 125, 160)
LMI.df <- data.frame(Year, Name, MPH, MPS = MPH * .44704)

NS <- c(15, 16, 9, 19, 19, 19, 14, 8, 11, 15, 17, 15, 18)
MH <- c(2, 5, 2, 5, 4, 2, 0, 2, 2, 4, 6, 2, 3)
Year <- 2007:2019

Counts.df <- data.frame(Year, NS, MH)
# sum(MH)/sum(NS) * 100
```

We note that `r round(sum(MH)/sum(NS) * 100, 0)`% of them had wind speeds exceeding 48.9 m/s. An increase of 5 percentage points over the earlier period examined in @ElsnerKossinJagger2008.

```{r}
suppressPackageStartupMessages(library(dplyr))
LMI.df %>%
  arrange(desc(MPS))
```

Plot as an ordered bar plot.
```{r}
LMI.df <- LMI.df %>%
  mutate(NAME = paste(Year, Name))

suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(forcats))

ggplot(data = LMI.df) +
  geom_segment(aes(x = reorder(NAME, MPS), y = 50,
                   xend = reorder(NAME, MPS), yend = MPS, 
                   color = MPS), size = 2.5) +
  coord_flip() + 
  scale_y_continuous(position = 'right') +
  xlab("") + ylab("Lifetime Highest Intensity (m/s)") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none")
```

```{r}
percentile <- (cumsum(table(LMI.df$MPS)) + (sum(NS) - sum(MH)))/sum(NS)
percentile.df <- as.data.frame(percentile)
percentile.df$MPS <- as.numeric(rownames(percentile.df))

( p85ws <- round(approx(x = percentile.df$percentile, y = percentile.df$MPS, xout = .85)$y, 1) )
```

Further we note that the 85th quantile wind speed has increased from 48.9 m/s to `r p85ws` m/s which is an increase of `r p85ws - 48.9` m/s and which matches an extrapolation that takes the trend of .63 m/s/yr from Table 1 in @ElsnerKossinJagger2008 and multiplies it by 7 (mid point number of years in a 13-year period). [48.9 + .63 * 7 = 53.3 m/s].

```{r}
p95 <- round(approx(x = percentile.df$MPS, y = percentile.df$percentile, xout = 60.3)$y, 1)
p95ws <- round(approx(x = percentile.df$percentile, y = percentile.df$MPS, xout = .95)$y, 1)
```

Furthermore we note that 5% of the storms exceeded 60.3 m/s over the earlier period (Table 1, 95th quantile) but has increased to `r (1 - p95) * 100`% over the later period. The 95th quantile wind speed has increased from 60.3 m/s to `r p95ws` m/s, which is somewhat stronger than the extrapolated trend [60.3 + .81 * 7 = 66 m/s].

It is hard to argue against this straightforward post publication analysis and the results raise the question of why @ElsnerKossinJagger2008 was largely ignored or discounted when writing "state-of-the-knowledge" reports on hurricanes and climate change (e.g., @KnutsonEtAl2019).

### Repeat for Western Pacific typhoons

Get data from IBTraCs http://ibtracs.unca.edu/index.php?name=browse-year-basin

```{r}
Year <- c(2007, 2007, 2007, 2007, 2007, 2007, 2007, 2007, 2007,
          2008, 2008, 2008, 2008, 2008, 2008, 2008, 2008,
          2009, 2009, 2009, 2009, 2009, 2009, 2009,
          2010, 2010, 2010, 2010,
          2011, 2011, 2011, 2011, 2011, 2011, 2011) 
Name <- c("Kong-Rey", "Yutu", "Man-Yi", "Usagi", "Sepat", "Nari", "Wipha", "Krosa", "Kijiki", 
          "Neoguri", "Rammansun", "Nakri", "Fengshen", "Nuri", "Sinlaku", "Hagupit", "Jangmi",
          "Kujira", "Vamco", "Choi-Wan", "Parma", "Melor", "Lupit", "Nida",
          "Kompasu", "Fanapi", "Megi", "Chaba",
          "Songda", "Ma-on", "Muifa", "Nanmadol", "Roke", "Nesat", "Nalgae")

KNT <- c(100, 130, 125, 120, 140, 125, 135, 130, 115,
         100, 135, 125, 110, 100, 125, 125, 140,
         115, 120, 140, 135, 150, 140, 155,
         105, 105, 160, 115,
         140, 115, 140, 140, 115, 115, 130)
          
```



## References
