---
title: The Increasing Intensity of Strong Tropical Cyclones
subtitle: Validating Elsner et al. (2008)
author: James B. Elsner
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
bibliography: References.bib
editor_options: 
  chunk_output_type: console
---

Literature:
https://www.nature.com/articles/s41467-019-08471-z
https://journals.ametsoc.org/doi/full/10.1175/JCLI-D-17-0898.1?journalCode=clim

Science gets validated through predictions. Using satellite derived wind speed estimates from tropical cyclones @ElsnerKossinJagger2008 showed that the strongest hurricanes were getting stronger through the period 1981-2006. In particular, they found that 15% (85th quantile) of all named storms across the North Atlantic had wind speeds exceeding 48.9 m/s (Table 1 in @ElsnerKossinJagger2008 ) with an upward trend of .63 m/s/yr in this quantile wind speed. They related this increase to a corresponding rise in ocean temperatures consistent with theory [@Emanuel1988]. The oceans have continued to warm since that paper was published so we would predict that the upward trend in the intensity of the strongest hurricanes has continued. 

Load the functions.
```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(quantreg)
```

### Start with hurricanes in the North Atlantic

Import the IBTrACS version 4 data: 
```{r}
NA.df <- read.csv(file = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/csv/ibtracs.NA.list.v04r00.csv",
                  header = TRUE,
                  stringsAsFactors = FALSE,
                  na.strings = " ")

NA.df <- NA.df %>%
  filter(SEASON != "Year") %>%
  mutate(Season = as.integer(SEASON)) %>%
  filter(Season >= 1981) %>%
  mutate(Time = as.Date(ISO_TIME),
         WindKt = as.numeric(USA_WIND),
         WindMS = WindKt * .5144447) %>%
  select(SID, Season, NUMBER, BASIN, NAME, ISO_TIME, Time, WindKt, WindMS)

write.csv(NA.df, file = "NA.csv")
```

Get wind speed quantiles for tropical cyclones with winds at least 33 m/s. First get per storm fastest wind speed.
```{r}
PerStormMax.df <- NA.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .9, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df))$coefficients

Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n())

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)

```

Plot a slope graph. Original code from https://acaird.github.io/computers/r/2013/11/27/slopegraphs-ggplot Data Source: IBTrACS v4 (Knapp et al. 2018)
```{r}
months <- 24
Early <- round(as.vector(unlist(Change.df[1, 2:4], use.names = FALSE)), 1)
Late <- round(as.vector(unlist(Change.df[2, 2:4], use.names = FALSE)), 1)
Group <- c("75th", "90th", "95th")
df <- data.frame(Early, Late, Group)

#l11 <- paste(df$Group, paste(df$Early, "m/s", sep = " "), sep = "\n")
l11 <- paste(df$Early, "m/s", sep = " ")
#l12 <- paste(df$Group, paste(df$Late, "m/s", sep = " "), sep = "\n")
l12 <- paste(df$Late, "m/s", sep = " ")
pNA <- ggplot(df) + 
  geom_segment(aes(x = 0, xend = months, y = Early, yend = Late, color = Group), size = .75, show.legend = FALSE) +
  scale_color_manual(values=c("gray50", "gray30", "black")) +
  theme(panel.background = element_blank()) +
  theme(panel.grid=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(axis.text=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("") + ylab("") +
  xlim((0-12), (months + 12)) +
  ylim(55, 87) +
  geom_text(label = l12, y = df$Late, x = rep.int(months, ncol(df)), hjust = -.2, size = 3) +
  geom_text(label = l11, y = df$Early, x = rep.int(0, ncol(df)), hjust = 1.2, size = 3) +
  geom_text(label = paste0("1981-2006\n(n = ", Change.df[1, 5], ")"), x = 0, y = 85, hjust = 1.2, size = 4) +
  geom_text(label = paste0("2007-2019\n(n = ", Change.df[2, 5], ")"), x = months, y = 85, hjust = 0, size = 4) 

( pNA <- pNA + ggtitle("North Atlantic Hurricanes") )
#  ggtitle("Percentile Wind Speeds", subtitle = "North Atlantic Ocean hurricanes [Data Source: IBTrACS v4 (Knapp et al. 2018)]")
```

Plot annual counts and test for significant difference in the number of hurricanes.
```{r}
AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df,
            exact = FALSE)
```

### Repeat for typhoons in the Western North Pacific
```{r}
WP.df <- read.csv(file = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/csv/ibtracs.WP.list.v04r00.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   na.strings = " ")

WP.df <- WP.df %>%
  filter(SEASON != "Year") %>%
  mutate(Season = as.integer(SEASON)) %>%
  filter(Season >= 1981) %>%
  mutate(Time = as.Date(ISO_TIME),
         WindKt = as.numeric(USA_WIND),
         WindMS = WindKt * .5144447) %>%
  select(SID, Season, NUMBER, BASIN, NAME, ISO_TIME, Time, WindKt, WindMS)

write.csv(WP.df, file = "WP.csv")
```

```{r}
PerStormMax.df <- WP.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .9, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df))$coefficients

Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n())

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)

months <- 24
Early <- round(as.vector(unlist(Change.df[1, 2:4], use.names = FALSE)), 1)
Late <- round(as.vector(unlist(Change.df[2, 2:4], use.names = FALSE)), 1)
Group <- c("75th", "90th", "95th")
df <- data.frame(Early, Late, Group)

#l11 <- paste(df$Group, paste(df$Early, "m/s", sep = " "), sep = "\n")
#l12 <- paste(df$Group, paste(df$Late, "m/s", sep = " "), sep = "\n")
l11 <- paste(df$Early, "m/s", sep = " ")
l12 <- paste(df$Late, "m/s", sep = " ")

pWP <- ggplot(df) + 
  geom_segment(aes(x = 0, xend = months, y = Early, yend = Late, color = Group), size = .75, show.legend = FALSE) +
  scale_color_manual(values=c("gray50", "gray30", "black")) +
  theme(panel.background = element_blank()) +
  theme(panel.grid=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(axis.text=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("") + ylab("") +
  xlim((0-12), (months + 12)) +
  ylim(55, 87) +
  geom_text(label = l12, y = df$Late, x = rep.int(months, ncol(df)), hjust = -.2, size = 3) +
  geom_text(label = l11, y = df$Early, x = rep.int(0, ncol(df)), hjust = 1.2, size = 3) +
  geom_text(label = paste0("1981-2006\n(n = ", Change.df[1, 5], ")"), x = 0, y = 85, hjust = 1.2, size = 4) +
  geom_text(label = paste0("2007-2019\n(n = ", Change.df[2, 5], ")"), x = months, y = 85, hjust = 0, size = 4) 

( pWP <- pWP + ggtitle("Western North Pacific Typhoons") )
```

Plot annual counts and test for significant difference in the number of hurricanes.
```{r}
AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df, 
            exact = FALSE)
t.test(NH ~ EarlyLate, 
       data = AnnualCount.df)
```

### Repeat for hurricanes in the Eastern North Pacific
```{r}
EP.df <- read.csv(file = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/csv/ibtracs.EP.list.v04r00.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   na.strings = " ")

EP.df <- EP.df %>%
  filter(SEASON != "Year") %>%
  mutate(Season = as.integer(SEASON)) %>%
  filter(Season >= 1981) %>%
  mutate(Time = as.Date(ISO_TIME),
         WindKt = as.numeric(USA_WIND),
         WindMS = WindKt * .5144447) %>%
  select(SID, Season, NUMBER, BASIN, NAME, ISO_TIME, Time, WindKt, WindMS)

write.csv(EP.df, file = "EP.csv")
```

```{r}
PerStormMax.df <- EP.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .9, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df))$coefficients


Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n())

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)

months <- 24
Early <- round(as.vector(unlist(Change.df[1, 2:4], use.names = FALSE)), 1)
Late <- round(as.vector(unlist(Change.df[2, 2:4], use.names = FALSE)), 1)
Group <- c("75th", "90th", "95th")
df <- data.frame(Early, Late, Group)

#l11 <- paste(df$Group, paste(df$Early, "m/s", sep = " "), sep = "\n")
#l12 <- paste(df$Group, paste(df$Late, "m/s", sep = " "), sep = "\n")
l11 <- paste(df$Early, "m/s", sep = " ")
l12 <- paste(df$Late, "m/s", sep = " ")
pEP <- ggplot(df) + 
  geom_segment(aes(x = 0, xend = months, y = Early, yend = Late, color = Group), size = .75, show.legend = FALSE) +
  scale_color_manual(values=c("gray50", "gray30", "black")) +
  theme(panel.background = element_blank()) +
  theme(panel.grid=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(axis.text=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("") + ylab("") +
  xlim((0-12), (months + 12)) +
  ylim(55, 87) +
  geom_text(label = l12, y = df$Late, x = rep.int(months, ncol(df)), hjust = -.2, size = 3) +
  geom_text(label = l11, y = df$Early, x = rep.int(0, ncol(df)), hjust = 1.2, size = 3) +
  geom_text(label = paste0("1981-2006\n(n = ", Change.df[1, 5], ")"), x = 0, y = 85, hjust = 1.2, size = 4) +
  geom_text(label = paste0("2007-2019\n(n = ", Change.df[2, 5], ")"), x = months, y = 85, hjust = 0, size = 4) 

( pEP <- pEP + ggtitle("Eastern North Pacific Hurricanes") )
```

Plot annual counts and test for significant difference in the number of hurricanes.
```{r}
AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df, 
            exact = FALSE)
```

### Repeat for Southern Indian
```{r}
SI.df <- read.csv(file = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/csv/ibtracs.SI.list.v04r00.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   na.strings = " ")

SI.df <- SI.df %>%
  filter(SEASON != "Year") %>%
  mutate(Season = as.integer(SEASON)) %>%
  filter(Season >= 1981 & Season < 2020) %>%
  mutate(Time = as.Date(ISO_TIME),
         WindKt = as.numeric(USA_WIND),
         WindMS = WindKt * .5144447) %>%
  select(SID, Season, NUMBER, BASIN, NAME, ISO_TIME, Time, WindKt, WindMS)

write.csv(SI.df, file = "SI.csv")
```

```{r}
PerStormMax.df <- SI.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .9, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df))$coefficients

Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n()) 

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)

months <- 24
Early <- round(as.vector(unlist(Change.df[1, 2:4], use.names = FALSE)), 1)
Late <- round(as.vector(unlist(Change.df[2, 2:4], use.names = FALSE)), 1)
Group <- c("75th", "90th", "95th")
df <- data.frame(Early, Late, Group)

#l11 <- paste(df$Group, paste(df$Early, "m/s", sep = " "), sep = "\n")
#l12 <- paste(df$Group, paste(df$Late, "m/s", sep = " "), sep = "\n")
l11 <- paste(df$Early, "m/s", sep = " ")
l12 <- paste(df$Late, "m/s", sep = " ")
pSI <- ggplot(df) + 
  geom_segment(aes(x = 0, xend = months, y = Early, yend = Late, color = Group), size = .75, show.legend = FALSE) +
  scale_color_manual(values=c("gray50", "gray30", "black")) +
  theme(panel.background = element_blank()) +
  theme(panel.grid=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(axis.text=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("") + ylab("") +
  xlim((0-12), (months + 12)) +
  ylim(55, 87) +
  geom_text(label = l12, y = df$Late, x = rep.int(months, ncol(df)), hjust = -.2, size = 3) +
  geom_text(label = l11, y = df$Early, x = rep.int(0, ncol(df)), hjust = 1.2, size = 3) +
  geom_text(label = paste0("1981-2006\n(n = ", Change.df[1, 5], ")"), x = 0, y = 85, hjust = 1.2, size = 4) +
  geom_text(label = paste0("2007-2019\n(n = ", Change.df[2, 5], ")"), x = months, y = 85, hjust = 0, size = 4) 

( pSI <- pSI + ggtitle("Southern Indian Cyclones") )
```

Plot annual counts and test for significant difference in the number of hurricanes.
```{r}
AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df, 
            exact = FALSE)
```

### Repeat for South Pacific
```{r}
SP.df <- read.csv(file = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/csv/ibtracs.SP.list.v04r00.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   na.strings = " ")

SP.df <- SP.df %>%
  filter(SEASON != "Year") %>%
  mutate(Season = as.integer(SEASON)) %>%
  filter(Season >= 1981 & Season < 2020) %>%
  mutate(Time = as.Date(ISO_TIME),
         WindKt = as.numeric(USA_WIND),
         WindMS = WindKt * .5144447) %>%
  select(SID, Season, NUMBER, BASIN, NAME, ISO_TIME, Time, WindKt, WindMS)

write.csv(SP.df, file = "SP.csv")
```

```{r}
PerStormMax.df <- SP.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .9, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df))$coefficients

Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n())

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)


months <- 24
Early <- round(as.vector(unlist(Change.df[1, 2:4], use.names = FALSE)), 1)
Late <- round(as.vector(unlist(Change.df[2, 2:4], use.names = FALSE)), 1)
Group <- c("75th", "90th", "95th")
df <- data.frame(Early, Late, Group)

#l11 <- paste(df$Group, paste(df$Early, "m/s", sep = " "), sep = "\n")
#l12 <- paste(df$Group, paste(df$Late, "m/s", sep = " "), sep = "\n")
l11 <- paste(df$Early, "m/s", sep = " ")
l12 <- paste(df$Late, "m/s", sep = " ")
pSP <- ggplot(df) + 
  geom_segment(aes(x = 0, xend = months, y = Early, yend = Late, color = Group), size = .75, show.legend = FALSE) +
  scale_color_manual(values=c("gray50", "gray30", "black")) +
  theme(panel.background = element_blank()) +
  theme(panel.grid=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(axis.text=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("") + ylab("") +
  xlim((0-12), (months + 12)) +
  ylim(55, 87) +
  geom_text(label = l12, y = df$Late, x = rep.int(months, ncol(df)), hjust = -.2, size = 3) +
  geom_text(label = l11, y = df$Early, x = rep.int(0, ncol(df)), hjust = 1.2, size = 3) +
  geom_text(label = paste0("1981-2006\n(n = ", Change.df[1, 5], ")"), x = 0, y = 85, hjust = 1.2, size = 4) +
  geom_text(label = paste0("2007-2019\n(n = ", Change.df[2, 5], ")"), x = months, y = 85, hjust = 0, size = 4) 

( pSP <- pSP + ggtitle("South Pacific Cyclones") )
```

Plot annual counts and test for significant difference in the number of hurricanes.
```{r}
AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df, 
            exact = FALSE)
```

### Repeat for Northern Indian
```{r}
NI.df <- read.csv(file = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/csv/ibtracs.NI.list.v04r00.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   na.strings = " ")

NI.df <- NI.df %>%
  filter(SEASON != "Year") %>%
  mutate(Season = as.integer(SEASON)) %>%
  filter(Season >= 1981) %>%
  mutate(Time = as.Date(ISO_TIME),
         WindKt = as.numeric(USA_WIND),
         WindMS = WindKt * .5144447) %>%
  select(SID, Season, NUMBER, BASIN, NAME, ISO_TIME, Time, WindKt, WindMS)

write.csv(NI.df, file = "NI.csv")
```

```{r}
PerStormMax.df <- NI.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .9, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .9, data = LMI.df))$coefficients

Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n())

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)


months <- 24
Early <- round(as.vector(unlist(Change.df[1, 2:4], use.names = FALSE)), 1)
Late <- round(as.vector(unlist(Change.df[2, 2:4], use.names = FALSE)), 1)
Group <- c("75th", "90th", "95th")
df <- data.frame(Early, Late, Group)

#l11 <- paste(df$Group, paste(df$Early, "m/s", sep = " "), sep = "\n")
#l12 <- paste(df$Group, paste(df$Late, "m/s", sep = " "), sep = "\n")
l11 <- paste(df$Early, "m/s", sep = " ")
l12 <- paste(df$Late, "m/s", sep = " ")
pNI <- ggplot(df) + 
  geom_segment(aes(x = 0, xend = months, y = Early, yend = Late, color = Group), size = .75, show.legend = FALSE) +
  scale_color_manual(values=c("gray50", "gray30", "black")) +
  theme(panel.background = element_blank()) +
  theme(panel.grid=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(axis.text=element_blank()) +
  theme(panel.border=element_blank()) +
  xlab("") + ylab("") +
  xlim((0-12), (months + 12)) +
  ylim(55, 87) +
  geom_text(label = l12, y = df$Late, x = rep.int(months, ncol(df)), hjust = -.2, size = 3) +
  geom_text(label = l11, y = df$Early, x = rep.int(0, ncol(df)), hjust = 1.2, size = 3) +
  geom_text(label = paste0("1981-2006\n(n = ", Change.df[1, 5], ")"), x = 0, y = 85, hjust = 1.2, size = 4) +
  geom_text(label = paste0("2007-2019\n(n = ", Change.df[2, 5], ")"), x = months, y = 85, hjust = 0, size = 4) 

( pNI <- pNI + ggtitle("Northern Indian Cyclones") )
```

Plot annual counts and test for significant difference in the number of hurricanes.
```{r}
AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df, 
            exact = FALSE)
```

Plot together using {patchwork}. Use 9 by 7.5 on pdf landscape
```{r}
library(patchwork)

(pWP + pEP)/(pSI + pNA)/(pSP + pNI)
```

All six basins (globally)
```{r}
AllSix.df <- rbind(WP.df, EP.df, SI.df, NA.df, SP.df, NI.df)
  
PerStormMax.df <- AllSix.df %>%
  group_by(SID) %>%
  summarize(Year = first(Season),
            Name = first(NAME),
            KTS = max(WindKt, na.rm = TRUE),
            MPS = KTS * .5144447,
            MPH = KTS * 1.15078,
            EarlyLate = Year > 2006) 

LMI.df <- PerStormMax.df %>%
  filter(KTS >= 64)

summary(rq(MPS ~ Year, tau = .5, data = LMI.df[!LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .5, data = LMI.df[LMI.df$EarlyLate,]))$coefficients
summary(rq(MPS ~ Year, tau = .5, data = LMI.df))$coefficients

summary(rq(MPS ~ Year, tau = .5, data = LMI.df[LMI.df$Year >= 2007,]))$coefficients
summary(rq(MPS ~ Year, tau = .5, data = LMI.df[LMI.df$Year >= 1999,]))$coefficients

quantile(LMI.df$MPS[LMI.df$Year < 1999], probs = .9)
quantile(LMI.df$MPS[LMI.df$Year >= 1999 & LMI.df$Year <= 2017], probs = .9)

quantile(LMI.df$MPS[LMI.df$Year >= 2007 & LMI.df$Year <= 2013], probs = .5)
quantile(LMI.df$MPS[LMI.df$Year >= 2013], probs = .5)

quantile(LMI.df$MPS[LMI.df$Year >= 2007 & LMI.df$Year <= 2013], probs = .9)
quantile(LMI.df$MPS[LMI.df$Year >= 2013], probs = .9)



Change.df <- LMI.df %>%
  group_by(EarlyLate) %>%
  summarize(q75 = quantile(MPS, probs = .75),
            q90 = quantile(MPS, probs = .9),
            q95 = quantile(MPS, probs = .95),
            nH = n())

Change.df %>%
  select(-nH) %>%
  pivot_longer(cols = c(q75, q90, q95)) %>%
  pivot_wider(values_from = value,
              names_from = EarlyLate) %>%
  group_by(name) %>%
  mutate(pChange = (`TRUE` - `FALSE`)/`TRUE` * 100)

AnnualCount.df <- PerStormMax.df %>%
  filter(KTS > 64) %>%
  group_by(Year) %>%
  summarize(NH = sum(KTS >= 64),
            MH = sum(KTS >= 96),
            EarlyLate = first(EarlyLate))

ggplot(data = AnnualCount.df, aes(x = Year, y = NH)) +
  geom_point() + geom_line() +
  geom_smooth(method = lm)

wilcox.test(NH ~ EarlyLate, 
            data = AnnualCount.df, 
            exact = FALSE)
```

### References
