---
title: "Stronger Hurricanes"
author: "James B. Elsner"
date: September 11, 2019
output: html_document
---

I think we should resist the urge to be critical of other work. Instead, I propose the best way to highlight our earlier work is to explain how it maps onto newer data. What does our analyses/model say about the set of the most recent hurricanes? How well does extrapolation or prediction do on these new data?

For example, using satellite derived wind speed estimates over the period 1981-2006 [Elsner et al. (2008)](http://myweb.fsu.edu/jelsner/PDF/Research/ElsnerKossinJagger2008.pdf) show that the strongest hurricanes are getting stronger. They relate this increase to rising ocean temperatures consistent with [theory](https://www.nature.com/articles/nature03906).

Since the oceans have continued to warm we would expect the upward trends to continue. In particular, for hurricanes occurring over the North Atlantic they find that 15% (85th quantile) of all named storms had wind speeds exceeding 48.9 m/s (Table 1) and an upward trend of .63 m/s/yr in this quantile wind speed.

Now considering all North Atlantic named storms (tropical storms and hurricanes) over the 13-year period 2007-2019 [data taken from Wikipedia] we note that 20% of them had wind speeds exceeding 48.9 m/s. An increase of 5 percentage points.
```{r}
Year <- c(2007, 2007, 2008, 2008, 2008, 2008, 2008, 2009, 2009, 2010, 2010, 2010, 2010, 2010,
          2011, 2011, 2011, 2011, 2012, 2012, 2014, 2014, 2015, 2015, 2016, 2016, 2016, 2016,
          2017, 2017, 2017, 2017, 2017, 2017, 2018, 2018, 2019)
Name <- c("Dean", "Felix", "Bertha", "Gustav", "Ike", "Omar", "Paloma", "Bill", "Fred",
          "Danielle", "Earl", "Igor", "Julia", "Karl", "Irene", "Katia", "Ophelia", "Rina",
          "Michael", "Sandy", "Edouard", "Gonzalo", "Danny", "Joaquin", "Gaston", "Matthew",
          "Nicole", "Otto", "Harvey", "Irma", "Jose", "Lee", "Maria", "Ophelia", "Florence",
          "Michael", "Dorian")
MPH <- c(175, 175, 125, 150, 145, 130, 140, 130, 120, 130, 145, 155, 140, 125, 120, 140, 140,
         115, 115, 115, 120, 145, 125, 155, 120, 165, 140, 115, 130, 180, 155, 115, 175, 115, 
         150, 160, 185)
LMI.df <- data.frame(Year, Name, MPH, MPS = MPH * .44704)

NS <- c(15, 16, 9, 19, 19, 19, 14, 8, 11, 15, 17, 15, 7)
MH <- c(2, 5, 2, 5, 4, 2, 0, 2, 2, 4, 6, 2, 1)
Year <- 2007:2019

Counts.df <- data.frame(Year, NS, MH)
sum(MH)/sum(NS) * 100
```

Further we note that the 85th quantile wind speed has increased from 48.9 m/s to 53.3 m/s which is an increase of 4.4 m/s and which matches precisely an extrapolation by taking the trend of .63 m/s/yr from Table 1 of Elsner et al. (2008) and multipling it by 7 (mid point of the 13-year period). [48.9 + .63 * 7 = 53.3 m/s]. You can't do much better than that.
```{r}
percentile <- (cumsum(table(LMI.df$MPS)) + (sum(NS) - sum(MH)))/sum(NS)
percentile.df <- as.data.frame(percentile)
percentile.df$MPS <- as.numeric(rownames(percentile.df))

approx(x = percentile.df$percentile, y = percentile.df$MPS, xout = .85)
```

Furthermore we note that only 5% of the storms exceeded 60.3 m/s over the earlier period (Table 1, 95th quantile) but that has increased to 10% (1 - .90) and the 95th quantile wind speed has increased from 60.3 m/s to 67.7 m/s consistent with extrapolating the trend [60.3 + .81 * 7 = 66 m/s].
```{r}
approx(x = percentile.df$MPS, y = percentile.df$percentile, xout = 60.3)
approx(x = percentile.df$percentile, y = percentile.df$MPS, xout = .95)
```

It is very hard to argue against this simple post publication analysis. And it implies that folks were negligent in writing "state-of-the-knowledge" reports on hurricanes and climate change.